---
title: "MSDS 6306 401 Case Study 1"
author: "Yao Yao"
date: "March 4, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Introduction: 



import libraries to i
```{r}
#require library(downloader)
#require library(ggplot2)
#require library(reshape2)
```

==Gather GDP and Educational data==

directory set up
```{r}
#setwd(".\\Data")
getwd()
list.files()
```

download files, read files into csv
```{r}
library(downloader)
download('https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv', destfile='GDPWeb.csv')
download('https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv', destfile='EducationalWeb.csv')
list.files() #make sure the files are there
```

==clean GDP Data==

import csv into raw GDP file and examine attributes
```{r}
GDPRaw <- read.csv('GDPWeb.csv',stringsAsFactors = FALSE, header = TRUE)
str(GDPRaw)
```

eliminate header rows, rows without GDP, and unused columns, reset column count, examine data
```{r}
GDPData <- GDPRaw[5:194,1:5]
rownames(GDPData) <- seq(length=nrow(GDPData))
head(GDPData)
tail(GDPData)
```
retitle columns and remove unused column
```{r}
colnames(GDPData) <- c("CountryCode","Ranking","x","Economy","US Dollars (millions)")
head(GDPData)
GDPData <- GDPData[,c("CountryCode","Ranking","Economy","US Dollars (millions)")]
head(GDPData)
```

set the ranking as integer and GDP as numeric for later analysis, recheck attributes
```{r}
GDPData$Ranking <- as.integer(GDPData$Ranking)
GDPData$`US Dollars (millions)` <- as.numeric(gsub(",", "", GDPData$`US Dollars (millions)`))
str(GDPData)
dim(GDPData)
```

==Clean Educational Data==

import raw education data from csv and examine attributes
```{r}
EducationRaw <- read.csv('EducationalWeb.csv',stringsAsFactors = FALSE, header = TRUE)
str(EducationRaw)
dim(EducationRaw)
```

==Merge Education and GDP data==

merge columns based on all rows, save into raw merge file, examine attributes
```{r}
MergeData1 <- merge(x = GDPData, y = EducationRaw, by ='CountryCode', all=TRUE)
head(MergeData1)
tail(MergeData1)
dim(MergeData1)
```

remove merged columns based on lack of value in GDP, country, or GDP ranking
examine attributes and reset column count based on country code
```{r}
MergeData2<-MergeData1[rowSums(is.na(MergeData1[,2:5]))==FALSE,]
head(MergeData2)
tail(MergeData2)
dim(MergeData2)
rownames(MergeData2) <- seq(length=nrow(MergeData2))
```

export data frames into csv file in the data directory
```{r}
write.csv(MergeData1, "MergeData1.csv")
write.csv(MergeData2, "MergeData2.csv")
write.csv(GDPRaw, "GDPRaw.csv")
write.csv(GDPData, "GDPData.csv")
write.csv(EducationRaw, "EducationRaw.csv")
write.csv(GDPData, "GDPData.csv")
```

==analysis to answer questions==

extract the number of rows with NA GDPs and number of matched rows
```{r}
print(paste0("Total Number of Rows in Merged Data: ", nrow(MergeData1)))
Matches<-sum(is.na(MergeData1$`US Dollars (millions)`) == FALSE & is.na(MergeData1$Income.Group) == FALSE)
print(paste0("Number of Matched Rows: ", Matches))
NARanking<-sum(is.na(MergeData1$Ranking) == TRUE)
print(paste0("Number of N/A Rankings: ", NARanking))
NAEconomy<-sum(is.na(MergeData1$Economy) == TRUE)
print(paste0("Number of N/A Economies: ", NAEconomy))
NAGDP<-sum(is.na(MergeData1$`US Dollars (millions)`) == TRUE)
print(paste0("Number of N/A GDPs: ", NAGDP))
NAIncomeGroup<-sum(is.na(MergeData1$Income.Group) == TRUE)
print(paste0("Number of N/A Income Groups: ", NAIncomeGroup))
NATotal<-sum(is.na(MergeData1$Ranking) == TRUE| is.na(MergeData1$Economy) == TRUE | MergeData1$`US Dollars (millions)` == TRUE | is.na(MergeData1$Income.Group) == TRUE)
print(paste0("Total Number of Unmatched Rows: ", NATotal))
print(paste0("Number of Rows in Merged Data without N/A values: ", nrow(MergeData2)))

```

1) 189 countries has GDP values matching with educational data

rank the merged data by ascending GDP, examine attributes and export dataset
```{r}
NegGDP <- MergeData2[order(-MergeData2$Ranking),] 
head(NegGDP)
str(NegGDP)
write.csv(NegGDP, "NegGDP.csv")
```

find 13th country with the ascending GDP
```{r}
country13NegGDP<-NegGDP[13,3]
country13NegGDP
NegGDP[12:13,2:4]
```
2) country #13 from ascending GDP, but St. Kitts and Nevis and Grenada are tied at 12th
St. Kitts is 13th because it comes second in the alphabet after Grenada

Assign a subset of High Income OECD countries and find the mean of their GDP rank
```{r}
HIOECD <- MergeData2[ which(MergeData2$Income.Group=='High income: OECD'), ]
head(HIOECD)
write.csv(HIOECD, "HIOECD.csv")
AvgGDPRank<- mean(HIOECD$Ranking)
AvgGDPRank
```

Assign a subset of High Income NonOECD countries and find the mean of their GDP rank
```{r}
HINonOECD <- MergeData2[ which(MergeData2$Income.Group=='High income: nonOECD'), ]
head(HINonOECD)
write.csv(HINonOECD, "HINonOECD.csv")
AvgGDPRank<- mean(HINonOECD$Ranking)
AvgGDPRank
```
3) Average GDP rankings by income groups

Plot country code and GDP in US dollars with income groupings as specific colors
```{r}
library(ggplot2)
color.codes<-as.character(c("#3399FF", "#FF0000","#0000FF", "#00FF00", "#D95F02"))
ggplot(data = NegGDP, aes(y = `US Dollars (millions)`, x =Income.Group, colour = Income.Group))+ 
   geom_boxplot() + scale_y_log10() +
  scale_colour_manual(breaks = NegGDP$Income.Group,values = unique(as.character(color.codes)))
```
4) boxplot distribution of log(GDP dollars) per income group

separate the summary statistics of GDP by income groups
```{r}
tapply(NegGDP$`US Dollars (millions)`, NegGDP$Income.Group, summary)
```
5) the summary statistics of GDP by income groups

breaks into 5 separate quantile groups, increment of 20%, writes quantiles into csv
```{r}
Quantiles<-cut(MergeData2$Ranking, breaks=quantile(MergeData2$Ranking,seq(0, 1, 0.2)))
head(quantiles)
write.csv(Quantiles, "Quantiles.csv")
```
Makes a table of GDP ranking quantile groups versus Income Group
```{r}
library(reshape2)
table(MergeData2$Income.Group, Quantiles)
```

create a new table featuring GDP ranking, country, and Income group. examine attributes
```{r}
IncomeSub <- MergeData2[c(2,3,6)]
head(IncomeSub)
write.csv(IncomeSub, "IncomeSub.csv")
```

Find the subset of countries that are lower middle in income group and top 38 in GDP ranking
```{r}
LowerMiddleTop38 <- IncomeSub[which(IncomeSub$Ranking <= 38 & IncomeSub$Income.Group == "Lower middle income"),]
LowerMiddleTop38
write.csv(LowerMiddleTop38, "LowerMiddleTop38.csv")
nrow(LowerMiddleTop38)
```

report the number of counties
6) 5 lower income Countries with top 38 GDP

Conclusion:


